### 2025-09-17 paper/abstract.tex

\begin{abstract}
We present a proof-of-concept implementation and evaluation of an agent-based LLM framework for authoring
\emph{transparent text}, natural language articles which provide interactions allowing a reader to explore the
relationship to the underlying data. We evaluate our approach on a subset of SciGen, a dataset consisting of
tables from scientific articles and their corresponding descriptions.
\end{abstract}

>>>>

\begin{abstract}
We present an agent-based LLM framework for authoring \emph{transparent text}, web-based scholarly articles
which allow a reader to explore the relationship to the underlying data by hovering over fragments of text.
Our approach builds on recent developments in data provenance techniques for general-purpose programming
languages, and tasks LLM agents with identifying fragments of text which can be computed from data, including
numerical values computed by aggregations like sum and mean, comparatives and superlatives like ``better
than'' and ``largest'', and trend-adjectives like ``growing''. We evaluate our approach on a subset of SciGen,
a dataset consisting of tables from scientific articles and their corresponding descriptions, which we extend
with hand-generated counterfactual test cases for evaluating how machine-generated expressions generalise in
the presence of changes to the underlying data. Our results show that some state-of-the-art models are able to
synthesis compound expressions that generate data-driven text that agrees with gold solutions.
\end{abstract}

### 2025-09-17 paper/sec/authoring-workflow.tex

\begin{itemize}
    \item \textbf{SuggestionAgent}. Based on an LLM, the SuggestionAgent receives as input the initial text and the reference data, and returns the elements in the text that can be replaced with expressions in the Fluid language.
    \item \textbf{InterpretationAgent}. Also based on an LLM, the InterpretationAgent receives as input a selection (either obtained from the SuggestionAgent or manually defined by the author) and generates an interpretation of the text as a Fluid expression.
\end{itemize}

@@

\begin{figure}[h]
    \centering
    \begin{tcolorbox}[colback=gray!10, colframe=gray!50, boxrule=0.5pt, arc=2pt,
        left=6pt, right=6pt, top=4pt, bottom=4pt]
        \VerbatimInput[fontsize=\small, formatcom=\normalfont]{../system-prompt/system-prompt.txt}
    \end{tcolorbox}
    \vspace{-0.5em}
    \caption{System-Prompt defined for the Authoring Assistant}
    \label{fig:system-prompt}
\end{figure}

>>>>

\begin{itemize}
    \item \textbf{SuggestionAgent}. LLM-based agent which identifies text fragments potentially computable from data.
    \item \textbf{InterpretationAgent}. LLM-based agent which, for a given text fragment (provided by
    SuggestionAgent or by the author), attempts to synthesise a Fluid expression computing the target text.
\end{itemize}

@@

\begin{figure}[h]
    \centering
    \begin{tcolorbox}[colback=gray!10, colframe=gray!50, boxrule=0.5pt, arc=2pt,
        left=6pt, right=6pt, top=4pt, bottom=4pt]
        \VerbatimInput[fontsize=\small, formatcom=\normalfont]{../system-prompt/system-prompt.txt}
    \end{tcolorbox}
    \vspace{-0.5em}
    \caption{System Prompt defined for the Authoring Assistant}
    \label{fig:system-prompt}
\end{figure}

### 2025-09-17 paper/sec/introduction.tex
Missing SciGen citation.
Start on abstract.

\lstset{
    aboveskip=-7pt,
    belowskip=-8.5pt,
    xleftmargin=0pt,
    xrightmargin=0pt
}

@@

In principle transparent programming languages can also be used to create natural language which is bidirectly
linked to data, by computing key fragments of natural language from data. For example, using the language
presented in this paper,

@@

\subsection{Use cases}
Two potential scenarios for this sort of technology:

\begin{enumerate}
\item \textbf{Authoring transparent text.} Someone authoring content for an online article, wants to create text
linked to raw data (and derivative data such as charts or tabular summaries), so that the evidence base for
the claims made in the text can be explored \emph{in situ}, by interacting with the text.

\item \textbf{Interpreting text after the fact.} Someone reading textual claims derived from open data (e.g. a
scientific paper or climate report), wants to retroactively link the text to queries over the available data
and gradually ``rationally reconstruct'' the relationship between the claims in the paper and the evidence
base. Perhaps just to aid their own comprehension, or to provide some kind of justified peer review.
\end{enumerate}

\noindent Here we focus on the first one because the second one requires a certain amount of additional setup.

>>>>

@@

In principle transparent programming languages can also be used to create natural language which is bidirectly
linked to data, by computing key fragments of natural language from data. The following example, adapted from
\cite{moosavi21}, shows how the language presented in this paper can be used to achieve this:

@@

We envisage two possible use cases for transparent text:

\begin{enumerate}
\item \textbf{Authoring.} Someone authoring content for an online article, wants to create text
linked to raw data (and derivative data such as charts or tabular summaries), so that the evidence base for
the claims made in the text can be explored \emph{in situ}, by interacting with the text.

\item \textbf{Reading or reviewing.} Someone reading textual claims derived from open data (e.g. a
scientific paper or climate report), wants to retroactively link the text to queries over the available data
and gradually ``rationally reconstruct'' the relationship between the claims in the paper and the evidence
base. Perhaps just to aid their own comprehension, or to provide some kind of justified peer review.
\end{enumerate}
